 ---
title: "Classify Types"
---

Use the classification metrics to calculate the climate type for each grid cell, using the Koppen-Trewartha classification system.

## ***Climate Periods***
```{r}
#| label: periods
# Climatological Reference Period
crp = "crp"
crp_beg <- 1961
crp_end <- 1990
crp_lbl <- paste("CRP ", crp_beg, "-", crp_end, sep = "")
# Recent Climate Period
rcp = "rcp"
rcp_beg <- 2015
rcp_end <- 2024
rcp_lbl <- paste("RCP ", rcp_beg, "-", rcp_end, sep = "")
```

## ***Packages***
```{r}
#| label: packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(sf)
  library(arrow)
  library(gt)
  library(units)
})
```

## ***Directories***
```{r}
#| label: directories
dir_rf <- "_data/00_reference/"
dir_rd <- "_data/03_transformed_data/"
dir_wr <- "_data/04_classified_data/"
```

## ***Palettes***
```{r}
#| label: palettes
pth <- paste0(dir_rf,"tcc_group_palette.csv")
tcc_group_palette <- read_csv(pth, show_col_types = FALSE)
pth <- paste0(dir_rf,"tcc_type_palette.csv")
tcc_type_palette <- read_csv(pth, show_col_types = FALSE)
```

## ***Map Projections***
```{r}
#| label: projections
rbn_prj <- "ESRI:54030"
```

## ***Global Coastline***
```{r}
#| label: coastline
dir <- dir_rf
fil <- "global-coastline.gpkg"
pth <- paste(dir, fil, sep = "")
coastline_sf <- st_read(pth, quiet=TRUE)
```

## ***Classification metrics***
```{r}
#| label: metrics
pth <- paste(dir_rd, "crp_determinants.parquet", sep = "")
crp_metrics_df <- read_parquet(pth) 
pth <- paste(dir_rd, "rcp_determinants.parquet", sep = "")
rcp_metrics_df <- read_parquet(pth) 
```

## ***Functions***


```{r}
#| label: fn_classify_tcc
fn_classify_tcc <- function(df) {
  df <- df |>
    mutate(
      z1 = case_when(
        tmax < 10 ~ "F",
        ptot < athr ~ "B",
        tmin >= 18 ~ "A",
        tg10 >= 8 ~ "C",
        tg10 >= 4 ~ "D",
        tg10 >= 1 ~ "E",
        TRUE ~ "X"
      ),
      zn = ""
    )

  df <- df |>
    mutate(
      z2 = case_when(
        z1 == "A" & (pl60 <= 2) ~ "r",
        z1 == "A" ~ "w",
        z1 == "B" & (ptot < athr/2) ~ "W",
        z1 == "B" ~ "S",

        z1 == "C" & (wtot >= stot * 3) & (pmin < 30) & (ptot < 890) ~ "s",
        z1 == "C" & (stot >= wtot * 10) ~ "w",
        z1 == "C" ~ "f",

        z1 == "D" & (tmin > 0) ~ "o",
        z1 == "D" ~ "c",
        z1 == "F" & (tmax > 0) ~ "t",
        z1 == "F" ~ "i",
        TRUE ~ ""
      ),
      zn = paste(z1, z2, sep = "")
    )
  df <- df|> select(lon, lat, area, lsm, zn)
  return(df)
}
```

## ***Apply classification system***
```{r}
#| label: classify
df1 <- fn_classify_tcc(crp_metrics_df)
df2 <- fn_classify_tcc(rcp_metrics_df)
tcc_df <- df1 %>% 
  left_join(df2 , by = c("lon", "lat", "lsm")) |>
  rename(czn = zn.x, rzn = zn.y, area = area.x) |>
  select(lon, lat, lsm, area, czn, rzn) |>
  mutate(czn = factor(czn, levels = tcc_type_palette$typ)) |>
  mutate(rzn = factor(rzn, levels = tcc_type_palette$typ))
```

## ***Convert to spatial data frames***
```{r}
#| label: spatial
tcc_sf <- st_as_sf(tcc_df, coords = c("lon", "lat"), crs = 4326)
```

## ***Combine geometries***
```{r}
#| label: combine
tcc_cmb_sf <- tcc_sf |>
  filter(lsm > 0) |>
  summarise(
    geometry = st_combine(geometry),
    mkm = sum(area*lsm),
    .by = c(czn, rzn)
  ) |>
  mutate(cz1 = substr(czn, 1, 1)) |>
  mutate(rz1 = substr(rzn, 1, 1)) |>
  mutate(pct = round(mkm / sum(mkm) * 100, 3)) |>
  select(czn, cz1, rzn, rz1, mkm, pct, geometry) |>
  arrange(czn, rzn) 
tcc_cmb_sf <- st_set_crs(tcc_cmb_sf, 4326)
```

## ***Save Results***
```{r}
#| label: save
fil <- "tcc_classifications.gpkg"
dir <- dir_wr
pth <- paste0(dir,fil)
st_write(tcc_cmb_sf, pth, delete_dsn = TRUE, quiet = TRUE)
```

```{r}
#| label: fn_create_map
fn_create_map <- function(plot_data, palette, cp_lbl) {

  p <- ggplot(plot_data) +
    geom_sf(aes(color = type), size = 0.01) +
    scale_color_manual(values = palette, drop = FALSE) +
    geom_sf(
      data = coastline_sf,
      color = "black",
      linewidth = 0.1
    ) +
    coord_sf(crs = rbn_prj) +
    labs(
      title = "Koppen-Trewartha Climate Classification",
      subtitle = cp_lbl,
      color = "Types"
    ) +
    guides(
      color = guide_legend(
        override.aes = list(
          size = 5 # legend point size only
        )
      )
    )
    return(p)
}
```

```{r}
#| label: fn_build_summary_table
fn_build_summary_table <- function(df, cp) {
  tbl <- df |>
    gt(rowname_col = "typ" ) |>
    tab_stubhead(label = "Type")|>
    tab_header(
      title = paste("Classification Summary:",cp),
      subtitle = "Sorted by Percentage of Area"
    ) |>

    # 2) center body *and* summary cells for Area
    cols_align(align = "center", columns = mkm) |>
    # 1) center the column header
    tab_style(
      style = cell_text(align = "center"),
      locations = cells_column_labels(columns = mkm)
    ) |>

    cols_label(
      lbl = "Description",
      mkm = "Area (Mkm²)",
      pct = "%"
    ) |>

    fmt_number(columns = c(mkm), decimals = 1) |>
    fmt_number(columns = c(pct), decimals = 2) |>
    grand_summary_rows(
      columns = c(mkm, pct),
      fns = list("Total" = ~ sum(.)),
      fmt = list(~ fmt_number(.x, decimals = 1))
    ) 
}
```

## ***Summary Table***

```{r}
#| label: build_t1
crp_tbl_df <- tcc_cmb_sf |>
  st_drop_geometry() |>
  rename(typ = czn) |>
  select(typ, mkm, pct) |>
  summarise(
    mkm = sum(mkm),
    pct = sum(pct),
    .by = typ
  )|>
  arrange(typ)

df1 <- crp_tbl_df

df2 <- tcc_type_palette |> select(typ, lbl)

dfj <- df1 |>
   left_join(df2, by = "typ") |>
   select(typ,lbl,mkm,pct)

t1 <- fn_build_summary_table(dfj, crp_lbl)
```

```{r}
#| label: build_t2
rcp_tbl_df <- tcc_cmb_sf |>
  st_drop_geometry() |>
  rename(typ = rzn) |>
  select(typ, mkm, pct) |>
  summarise(
    mkm = sum(mkm),
    pct = sum(pct),
    .by = typ
  ) |>
  arrange(typ)

df1 <- rcp_tbl_df

df2 <- tcc_type_palette |> select(typ, lbl)

dfj <- df1 |>
   left_join(df2, by = "typ") |>
   select(typ,lbl,mkm,pct)

t2 <- fn_build_summary_table(dfj, rcp_lbl)
```
## ***Climate Group Visualization***

```{r}
#| label: v1
plot_data <- tcc_cmb_sf |>
  select(czn) |>
  rename(type = czn)

type_levels <- tcc_type_palette$typ

type_palette <- setNames(
  tcc_type_palette$hex,
  tcc_type_palette$typ
)

plot_data <- plot_data |>
  mutate(type = factor(type, levels = type_levels))

plot_palette <- type_palette[levels(plot_data$type)]

v1 <- fn_create_map(plot_data,plot_palette,crp_lbl)
```

```{r}
#| label: v2
plot_data <- tcc_cmb_sf |>
  select(rzn) |>
  rename(type = rzn)

type_levels <- tcc_type_palette$typ

type_palette <- setNames(
  tcc_type_palette$hex,
  tcc_type_palette$typ
)

plot_data <- plot_data |>
  mutate(type = factor(type, levels = type_levels))

plot_palette <- type_palette[levels(plot_data$type)]

v2 <- fn_create_map(plot_data,plot_palette,rcp_lbl)
```

```{r}
#| label: t1-v1
t1
```
```{r}
#| label: v1_hw
#| fig-width: 10
#| fig-aspect: .7   
#| out-width: 100%
v1 
```
```{r}
#| label: t2-v2
t2
```

```{r}
#| label: v2_hw
#| fig-width: 10
#| fig-aspectt: .7  
#| out-width: 100%
v2
```